Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders

### 2025-07-21 Paper/Paper.qmd
update paper

The systematic review was conducted in accordance with the above listed goals, the preregistration protocol, and the PRISMA guidelines. The methods are detailed below.

@@

A systematic search was conducted on four databases: PsycINFO, PubMed, Scopus, Web of Science. No time limits were set, however virtually no records from before 1990 were expected, as taxometric methods were not yet developed or diffuse then. The following search query was performed on titles, abstracts, and/or keywords:

@@

Where the wildcard "\*" was not supported (i.e., PubMed) it was replaced with full terms ("disorder", "disability" instead of "dis\*"; "mathematical", "mathematics", "math" instead of "math\*").

@@

Inclusion criteria:

-   The study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);

-   The target population is individuals with NDDs or related traits;

-   The study draws a conclusion about the latent structure (categorical vs. dimensional) or reports sufficient information for an independent reader to reach a conclusion based on those results (e.g., reports CCFIs).

Exclusion criteria:

-   Theoretical, methodological, or simulation articles without empirical data;

-   Articles not reporting any behavioral data (e.g., only neuroimaging or neurophysiological data);

-   Reviews and/or meta-analyses;

-   Studies using other latent structure methods only (e.g., LPA, mixture modeling) without any taxometric methods being employed.

@@

Two independent human coders (MC and RP) screened and coded all articles. All discrepancies were resolved through discussion, and later revised by the PI (ET). No AI tools were used in the process, except in occasional cases by the human coders to assist them clarify the content of articles.

Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataframe followed a long-form data structure, with some articles repeating over multiple rows. Coded variables are listed in the data dictionary table reported in the preregistration protocol, to which a few variables for notes and coders agreement were added.

@@

@@

We simulate a dimensional scenario that makes it visually compelling how artificial admixture of the first type mentioned above creates pseudotaxa. A single normally-distributed latent variable $X$ is generated, that defines the target condition. A series of observed indicators $x1, x2, etc.$ are simulated, incorporating the $X$ latent variable plus some random normally-distributed noise ($R^2 = 0.77$). Individuals above the 99^th^ percentile of $X$ receive a diagnosis via clinical assessment. A sample of $N = 1000$ individuals is collected from the general population. Since the base rate is low (on average, only about $n = 10$ cases are expected to naturally present a diagnosis), researchers select and recruit an additional $n = 1000$ cases with a clinical diagnosis from the general population, thus creating artificial admixture. All data simulation is available in the online materials at xxxxxxxxxxxxxxxx.

An instance of the simulation, plotting two observed variables for illustrative purposes, is shown in @FIGURE_xx. Color is added for further clarity to distinguish individuals with and without a diagnosis, but two clusters still clearly emerge. Indicator validity is good (about $Cohen's$ $d = 2$), while nuisance covariance is partly problematic: for the clinical cases $r \approx 0$, but for the nonclinical cases $r = 0.61$. It is worth reminding that no taxa exist here, and the appearance of a categorical structure is mainly imposed by admixture, as shown below.

A Monte Carlo simulation was then performed, with 1000 iterations. Taxometric analysis was conducted using the "RTaxometric" package of R. As recommended, CCFI was used for inference: CCFI varies between 0 (maximal evidence for dimensional structure) to 1 (maximal evidence for categorical structure), with 0.5 indicating inconclusive results. Performed procedures were, as per default: MAMBAC, MAXEIG, and L-Mode. A third $x3$ observed indicator (similar to those above) was entered to allow conducting all these taxometric procedures. Over 1000 simulated datasets, results always strongly favored a categorical structure: average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\] (average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx). The smallest combined CCFI over 1000 iterations was xxxx. To confirm that artificial admixture also leads to misleading results when performing clustering, Gaussian Mixture Models (GMM) was also run testing 1 vs 2 components, using the "mclust" package of R: over 1000 iterations, results always favored the 2-component relative to the 1-component model, with an average $\Delta BIC = xxxx$ in favor of the 2-component model.

To confirm that the categorical structure was incorrectly favored specifically because of artificial admixture, we run another Monte Carlo simulation like the one above, but without admixture. In this case, a single sample of $N = 2000$ observations was drawn at each iteration. Therefore, nuisance covariance remains strong, but there is no admixture. Average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\]; average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx. Largest combined CCFI was xxxx. Thus, results remain frequently around inconclusive, probably due to nuisance covariance, but never strongly favor a categorical structure. GMM also consistently favored the 1-component model (over the 2-component model), with an average $\Delta BIC = xxxx$ in favor of the 1-component model. Therefore, admixture can be an extremely strong driver of false categorical findings, especially when many clinical cases are separately recruited and added to the sample.

@@

The present review suggests that taxometric analysis has been underused in neurodevelopmental research, despite its unique relevance for the recent debate on the categorical vs dimensional nature of disorders [@astle2022trandiagnostic; @michelini2024world; @happe2021both]. Focusing on selected NNDs, we identified 14 relevant records. The underuse of taxometric analysis in this field was expected: @haslam2020metaanalysis identified 183 articles employing taxometric methods on psychopathology, of which only 15 concerned childhood (these included various NNDs but also other conditions such as sleep problems). What is most worrying, many of the articles that we reviewed present serious methodological limitations and/or incomplete reporting of relevant information.

>>>>

The systematic review was conducted according to the Preferred Reporting items for Systematic Reviews and Meta-Analysis 2020 guidelines. A review protocol was preregistered on OSF registries at the following link: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0 The methods are detailed below.

@@

A systematic search was conducted for relevant peer-reviewed articles in April 2025 using four electronic databases: PsycINFO, PubMed, Scopus, Web of Science. No time limits were set, although no records were expected from before 1990, as taxometric methods were not yet developed or diffuse then. The following search query, reflecting the methods and the target populations of interest, was performed on titles, abstracts, and keywords:

@@

in databases where the wildcard "\*" was not supported (i.e., PubMed), it was replaced with full terms ("disorder", "disability" instead of "dis\*"; "mathematical", "mathematics", "math" instead of "math\*").

@@

Inclusion criteria were: 1) the study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG); 2) the target population is individuals with NDDs or related traits; 3) the study draws a conclusion about the latent structure (categorical vs. dimensional) or reports sufficient information for an independent reader to reach a conclusion based on those results (e.g., reports CCFIs).

Exclusion criteria were: 1) the study only reports a theoretical, methodological, or simulation articles without empirical data; 2) the article does not report any behavioral data (e.g., only neuroimaging or neurophysiological data; this was later reconsidered, but no studies were excluded for this reason in any case); 3) the study only consists of a review and/or meta-analysis; 4) the study employs other latent structure methods only (e.g., LPA, mixture modeling), and not taxometric analysis. In addition, following the preregistration protocol, the search was limited to published, peer-reviewed journal articles, as a primary aim was to provide a critical review of the existing literature. However, one potentially interesting record excluded as non-journal article was separately discussed below in the Results section.

@@

Two independent human coders (MC and RP) screened and coded all articles. No AI tools were used in the process, except in occasional cases by the human coders to assist them clarify the content of articles. All discrepancies were resolved through discussion, and later revised by the PI (ET).

Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataframe followed a long-form data structure, with some articles repeating over multiple rows. Coded variables were listed in the data dictionary table defined in the preregistration protocol, to which a few variables for notes and coders agreement were added.

Inter-coder agreement was fair to good in all phases. For the title+abstract screening phase, Cohen's k = 0.79 (34 out of 38 matching independent decisions); for the full-text eligibility screening, Cohen's k = 0.52 (15 out of 19). For final data extraction, 84% (71 out of 85) rows were extracted by both coders, with good agreement on the crucial information: for the "authors' taxonic conclusion" agreement was 94%, for artificial admixture was 95% at the sample/subsample level.

@@

A total of 110 records were identified through database searches. After removing 67 duplicates and excluding 5 records (dissertations or book sections, only one of which appeared to include primary taxometric analyses, but could not be retrieved), 38 unique articles were screened. Nineteen were excluded at the title/abstract level. Full-text assessment was conducted on the remaining 19 reports, with 7 excluded for not meeting inclusion criteria (4 did not use taxometric analysis; 3 did not target neurodevelopmental disorders). An additional 4 records were identified through citation tracking and online browsing, of which 2 were excluded (1 methodological article; 1 not employing taxometric methods). In total, 14 studies were included in the review. The PRISMA flowchart in @fig-prisma summarizes the selection process.

@fig-prisma *about here*

![PRISMA 2020 flow diagram illustrating the study screening and selection process](Figures/PRISMA_flow_diagram.png){#fig-prisma}

## Additional Use of Cluster Analysis

Among the 14 identified articles, six also employed methods of cluster analysis (in all but one case using latent class or latent profile analysis). In five of the six cases, cluster analysis was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was performed as a secondary analysis [@frazier2010asd]. In all cases, clustering results were interpreted as supporting taxonic conclusions, always converging with those from taxometric analyses, with the exception of @frazier2007adhd, where clustering results were partially inconclusive yet suggested the presence of subgroups, in contrast to the dimensional structure indicated by taxometric analysis. In all but two studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.

@@

```{r}
load("../R/simulateAdmixtResults.RData")
```

We simulate a dimensional scenario that makes it visually compelling how artificial admixture of the first type mentioned above creates pseudotaxa. A single normally-distributed latent variable $X$ is generated, that defines the target condition. A series of observed indicators $x1, x2, etc.$ are simulated, incorporating the $X$ latent variable plus some random normally-distributed noise ($R^2 = 0.77$). Individuals above the 99^th^ percentile of $X$ receive a diagnosis via clinical assessment. A sample of $N = 1000$ individuals is collected from the general population. Since the base rate is low (on average, only about $n = 10$ cases are expected to naturally present a diagnosis), researchers select and recruit an additional $n = 1000$ cases with a clinical diagnosis from the general population, thus creating artificial admixture. All data simulation is available in the online materials at https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba

An instance of the simulation, plotting two observed variables for illustrative purposes, is shown in @fig-admixture. Color is added for further clarity to distinguish individuals with and without a diagnosis, but two clusters still clearly emerge. Indicator validity is good (about $Cohen's$ $d = 2$), while nuisance covariance is partly problematic: for the clinical cases $r \approx 0$, but for the nonclinical cases $r = 0.61$. It is worth reminding that no taxa exist here, and the appearance of a categorical structure is mainly imposed by admixture, as shown below.

@fig-admixture *about here*

![A simulated example of artificial admixture, where a large number of individuals with a diagnosis are separately recruited and added to the sample (total N = 2000). The true underlying structure is strictly dimensional.](Figures/figureAdmixture.png){#fig-admixture}

A Monte Carlo simulation was then performed, with 1000 iterations. Taxometric analysis was conducted using the "RTaxometric" package of R. As recommended, CCFI was used for inference: CCFI varies between 0 (maximal evidence for dimensional structure) to 1 (maximal evidence for categorical structure), with 0.5 indicating inconclusive results. Performed procedures were, as per default: MAMBAC, MAXEIG, and L-Mode. A third $x3$ observed indicator (similar to those above) was entered to allow conducting all these taxometric procedures. Over 1000 simulated datasets, results always strongly favored a categorical structure: median combined CCFI = `r formatC(median(resultsAdmix$CCFI.mean),2,format="f")`, 95% quantile interval \[`r formatC(quantile(resultsAdmix$CCFI.mean,.025),2,format="f")`, `r formatC(quantile(resultsAdmix$CCFI.mean,.975),2,format="f")`\]; median MAMBAC CCFI = `r formatC(median(resultsAdmix$CCFI.MAMBAC),2,format="f")`; average MAXEIG CCFI= `r formatC(median(resultsAdmix$CCFI.MAXEIG),2,format="f")`; average L-Mode CCFI = `r formatC(median(resultsAdmix$CCFI.LMode),2,format="f")`. The smallest combined CCFI over 1000 iterations was `r formatC(min(resultsAdmix$CCFI.mean),2,format="f")`. Therefore, admixture can be an extremely strong driver of false categorical findings, especially when many clinical cases are separately recruited and added to the sample.

To confirm that artificial admixture leads to misleading results also when performing clustering, Gaussian Mixture Models (GMM) was also run testing 1 vs 2 components, using the "mclust" package of R. GMM is a type of finite mixture model, similar to latent profile analysis, but more flexible in that it also models indicator covariances. Over 1000 iterations, results always favored the 2-component relative to the 1-component model, with a median $\Delta BIC$ = `r formatC(median(resultsAdmix$gmmBIC1-resultsAdmix$gmmBIC2),2,format="f")` in favor of the 2-component model.

To confirm that the categorical structure was incorrectly favored specifically because of artificial admixture, we run another Monte Carlo simulation like the one above, but without admixture. In this case, a single sample of $N = 2000$ observations was drawn at each iteration. Therefore, nuisance covariance remains strong, but there is no admixture. Median combined CCFI = `r formatC(median(resultsNoAdmix$CCFI.mean),2,format="f")`, 95% quantile interval \[`r formatC(quantile(resultsNoAdmix$CCFI.mean,.025),2,format="f")`, `r formatC(quantile(resultsNoAdmix$CCFI.mean,.975),2,format="f")`\]; median MAMBAC CCFI = `r formatC(median(resultsNoAdmix$CCFI.MAMBAC),2,format="f")`; average MAXEIG CCFI= `r formatC(median(resultsNoAdmix$CCFI.MAXEIG),2,format="f")`; average L-Mode CCFI = `r formatC(median(resultsNoAdmix$CCFI.LMode),2,format="f")`. Thus, results remain frequently around inconclusive, probably due to nuisance covariance, but do not risk strongly favoring a categorical structure. GMM also consistently favored the 1-component model (over the 2-component model), with an average $\Delta BIC$ = `r formatC(median(resultsNoAdmix$gmmBIC2-resultsNoAdmix$gmmBIC1),2,format="f")` in favor of the 1-component model.

@@

The present review suggests that taxometric analysis has been underused in neurodevelopmental research, and that many articles using it present relevant methodological limitations or incomplete reporting of crucial information. This is unfortunate in light of the unique relevance of taxometric analysis for the current debate on the categorical vs dimensional nature of NDDs[@astle2022trandiagnostic; @michelini2024world; @happe2021both]. The underuse of taxometric analysis in this field was expected: of the 183 articles employing taxometric methods on psychopathology identified by @haslam2020metaanalysis only 15 were placed in the "childhood" category (encompassing all types, even conditions such as sleep problems).
